Reanalysis Step 3: Comprehensive Imputation Diagnostics

### 2025-08-30 reanalysis-3-imputation_diagnostics.qmd
clean up diagnsotics

---
title: "Reanalysis Step 2b: Multiple Imputation Outcome Diagnostics"
subtitle: "Assessing plausibility of imputed visual acuity outcomes"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 3
    fig-width: 10
    fig-height: 6
---

@@

This notebook examines whether our multiple imputation procedure for visual acuity outcomes is producing clinically plausible values. We're particularly concerned about:

1. **Temporal consistency**: Do imputed 2-year outcomes respect the trajectory from baseline through intermediate timepoints?
2. **Clinical plausibility**: Are changes in vision realistic given what we know about disease progression?
3. **Distribution similarity**: Do imputed values have similar distributions to observed values?

@@

library(mice)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
library(knitr)

# Initialize logging
log_file <- init_log("imputation_diagnostics")
```

## Load Data

```{r load-data}
# Load the original data to identify which values were imputed
df_original <- readRDS(file.path(reanalysis_data_dir, "df_for_imputation.rds"))

# Load the imputed datasets
imputed_datasets <- readRDS(file.path(reanalysis_data_dir, "imputed_datasets.rds"))
n_imputations <- length(imputed_datasets)

log_message(paste0("Loaded ", n_imputations, " imputed datasets"))
log_message(paste0("Total observations: ", nrow(df_original)))
```

## Identify Imputed vs Observed Outcomes

```{r identify-imputed}
# In the original data, cohort_id tells us who had observed outcomes
# "submitted" = has true 2-year outcome
# "excluded" = lacks true 2-year outcome (these were imputed)

# Create more nuanced grouping
df_original <- df_original |>
  mutate(
    outcome_was_imputed = cohort_id == "excluded",
    had_intermediate = !is.na(intermediate_VA_logMAR),

    # Create detailed imputation groups
    imputation_group = case_when(
      !outcome_was_imputed ~ "Observed",
      outcome_was_imputed & had_intermediate ~ "Imputed (with intermediate)",
      outcome_was_imputed & !had_intermediate ~ "Imputed (no intermediate)",
      TRUE ~ "Unknown"
    ),
    imputation_group = factor(imputation_group, 
                             levels = c("Observed", 
                                       "Imputed (with intermediate)", 
                                       "Imputed (no intermediate)"))
  )

# Summary of imputation status
imputation_summary <- df_original |>
  group_by(imputation_group) |>
  summarise(
    n = n(),
    pct = round(100 * n / nrow(df_original), 1),
    had_baseline = sum(!is.na(baseline_VA_logMAR)),
    pct_baseline = round(100 * had_baseline / n, 1),
    .groups = "drop"
  )

kable(imputation_summary,
      col.names = c("Group", "N", "%", "Has Baseline", "% Baseline"),
      caption = "Detailed breakdown of imputation groups")

# Baseline characteristics by group
baseline_chars <- df_original |>
  group_by(imputation_group) |>
  summarise(
    # Demographics
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    pct_male = round(100 * mean(gender == "Male", na.rm = TRUE), 1),

    # Clinical characteristics - check if variables exist
    mean_CCI = if("CCI" %in% names(df_original)) mean(CCI, na.rm = TRUE) else NA_real_,
    sd_CCI = if("CCI" %in% names(df_original)) sd(CCI, na.rm = TRUE) else NA_real_,
    mean_DCSI = if("DCSI" %in% names(df_original)) mean(DCSI, na.rm = TRUE) else NA_real_,
    sd_DCSI = if("DCSI" %in% names(df_original)) sd(DCSI, na.rm = TRUE) else NA_real_,

    # Baseline vision
    mean_baseline_VA = mean(baseline_VA_logMAR, na.rm = TRUE),
    sd_baseline_VA = sd(baseline_VA_logMAR, na.rm = TRUE),

    # Treatment indicators
    pct_anti_VEGF = round(100 * mean(anti_VEGF == 1, na.rm = TRUE), 1),
    pct_PRP = round(100 * mean(PRP_flag == 1, na.rm = TRUE), 1),

    # Lapse status
    pct_lapsed = round(100 * mean(ever_lapse_binary == 1, na.rm = TRUE), 1),

@@

kable(baseline_chars |> 
        mutate(across(where(is.numeric), ~round(., 2))),
      caption = "Baseline characteristics by imputation group")
```

## Calculate Change Scores

```{r calculate-changes}
# Add row identifier to original data for proper merging
df_original <- df_original |>
  mutate(row_id = row_number())

# Function to calculate change scores for a single imputed dataset
calculate_change_scores <- function(imp_data, original_data) {
  imp_data |>
    mutate(
      row_id = row_number(),
      # Add imputation status from original data
      outcome_was_imputed = original_data$outcome_was_imputed,
      had_intermediate = original_data$had_intermediate,
      imputation_group = original_data$imputation_group,

      # Calculate changes
      change_baseline_to_outcome = outcome_VA_logMAR - baseline_VA_logMAR,
      change_baseline_to_intermediate = intermediate_VA_logMAR - baseline_VA_logMAR,
      change_intermediate_to_outcome = outcome_VA_logMAR - intermediate_VA_logMAR,

      # Categorize changes (negative = improvement, positive = worsening)
      # Using 0.1 logMAR as clinically meaningful change (approximately 1 line on eye chart)
      outcome_direction = case_when(
        change_baseline_to_outcome < -0.1 ~ "Improved",
        change_baseline_to_outcome > 0.1 ~ "Worsened",
        TRUE ~ "Stable"
      ),

      # Flag extreme changes (clinically implausible)
      extreme_improvement = change_baseline_to_outcome < -1.0,  # >10 lines improvement
      extreme_worsening = change_baseline_to_outcome > 1.5,      # >15 lines worsening

      # For those with intermediate values, check trajectory consistency
      trajectory_reversal = case_when(
        !is.na(change_baseline_to_intermediate) & !is.na(change_intermediate_to_outcome) ~
          sign(change_baseline_to_intermediate) != sign(change_intermediate_to_outcome),
        TRUE ~ FALSE
      )
    )
}

# Calculate for all imputed datasets
all_changes <- list()
for (i in 1:n_imputations) {
  all_changes[[i]] <- calculate_change_scores(imputed_datasets[[i]], df_original)
}

# For summary statistics, use the first imputation as representative
changes_df <- all_changes[[1]]
```

## Compare Change Distributions

### Summary Statistics

```{r summary-stats}
# Compare change scores across all three groups
change_summary_detailed <- changes_df |>
  group_by(imputation_group) |>
  summarise(
    n = n(),
    mean_change = mean(change_baseline_to_outcome, na.rm = TRUE),
    sd_change = sd(change_baseline_to_outcome, na.rm = TRUE),
    median_change = median(change_baseline_to_outcome, na.rm = TRUE),
    q25_change = quantile(change_baseline_to_outcome, 0.25, na.rm = TRUE),
    q75_change = quantile(change_baseline_to_outcome, 0.75, na.rm = TRUE),
    min_change = min(change_baseline_to_outcome, na.rm = TRUE),
    max_change = max(change_baseline_to_outcome, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    iqr_change = q75_change - q25_change
  )

kable(change_summary_detailed |> 
        select(imputation_group, n, mean_change, sd_change, median_change, iqr_change, min_change, max_change),
      digits = 3,
      col.names = c("Group", "N", "Mean", "SD", "Median", "IQR", "Min", "Max"),
      caption = "Distribution of vision changes (outcome - baseline) by detailed imputation group")

# Also keep simplified version for compatibility
change_summary <- changes_df |>
  group_by(outcome_was_imputed) |>
  summarise(
    n = n(),
    mean_change = mean(change_baseline_to_outcome, na.rm = TRUE),
    sd_change = sd(change_baseline_to_outcome, na.rm = TRUE),
    median_change = median(change_baseline_to_outcome, na.rm = TRUE),
    q25_change = quantile(change_baseline_to_outcome, 0.25, na.rm = TRUE),
    q75_change = quantile(change_baseline_to_outcome, 0.75, na.rm = TRUE),
    min_change = min(change_baseline_to_outcome, na.rm = TRUE),
    max_change = max(change_baseline_to_outcome, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    iqr_change = q75_change - q25_change,
    group = ifelse(outcome_was_imputed, "Imputed", "Observed")
  )

# Statistical tests across three groups
anova_test <- aov(change_baseline_to_outcome ~ imputation_group, data = changes_df)
anova_summary <- summary(anova_test)

# Pairwise comparisons
pairwise_tests <- pairwise.t.test(changes_df$change_baseline_to_outcome, 
                                  changes_df$imputation_group,
                                  p.adjust.method = "bonferroni")

# Format ANOVA results
anova_results <- data.frame(
  Test = "One-way ANOVA",
  F_Statistic = anova_summary[[1]]$`F value`[1],
  df1 = anova_summary[[1]]$Df[1],
  df2 = anova_summary[[1]]$Df[2],
  P_Value = anova_summary[[1]]$`Pr(>F)`[1],
  Interpretation = ifelse(anova_summary[[1]]$`Pr(>F)`[1] < 0.05, 
                         "Groups differ significantly", "No significant difference")
)

kable(anova_results, digits = 4, caption = "ANOVA comparing change distributions across three groups")

# Format pairwise results
pairwise_matrix <- pairwise_tests$p.value
pairwise_df <- data.frame(
  Comparison = c("Imputed (with intermediate) vs Observed",
                "Imputed (no intermediate) vs Observed",
                "Imputed (no intermediate) vs Imputed (with intermediate)"),
  P_Value = c(pairwise_matrix[1,1], pairwise_matrix[2,1], pairwise_matrix[2,2]),
  Significant = ifelse(c(pairwise_matrix[1,1], pairwise_matrix[2,1], pairwise_matrix[2,2]) < 0.05,
                      "Yes", "No")
)

kable(pairwise_df, digits = 4, caption = "Pairwise comparisons (Bonferroni adjusted)")

# Also do simplified tests for backward compatibility
t_test <- t.test(change_baseline_to_outcome ~ outcome_was_imputed, data = changes_df)
ks_test <- ks.test(
  changes_df$change_baseline_to_outcome[changes_df$outcome_was_imputed],
  changes_df$change_baseline_to_outcome[!changes_df$outcome_was_imputed]
)
```

### Visualization of Change Distributions

```{r change-distributions}
#| fig-height: 8
#| fig-width: 12

# Prepare data for plotting - using detailed groups
plot_data_detailed <- changes_df |>
  filter(!is.na(change_baseline_to_outcome))

# Also prepare simplified version for some plots
plot_data <- changes_df |>
  mutate(
    Group = ifelse(outcome_was_imputed, "Imputed", "Observed"),
    Group = factor(Group, levels = c("Observed", "Imputed"))
  ) |>
  filter(!is.na(change_baseline_to_outcome))

# 1. Density plots - detailed groups
p1_detailed <- ggplot(plot_data_detailed, aes(x = change_baseline_to_outcome, fill = imputation_group)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(data = change_summary_detailed, 
             aes(xintercept = mean_change, color = imputation_group),
             linetype = "solid", size = 1) +
  scale_fill_manual(values = c("Observed" = "#2E86AB", 
                              "Imputed (with intermediate)" = "#A23B72",
                              "Imputed (no intermediate)" = "#F18F01")) +
  scale_color_manual(values = c("Observed" = "#2E86AB", 
                               "Imputed (with intermediate)" = "#A23B72",
                               "Imputed (no intermediate)" = "#F18F01")) +
  labs(
    title = "Distribution of Vision Changes by Detailed Group",
    subtitle = "Vertical lines show group means",
    x = "Change in logMAR (positive = worsening)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# 1b. Simplified density plot (for comparison)
p1 <- ggplot(plot_data, aes(x = change_baseline_to_outcome, fill = Group)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(data = change_summary, 
             aes(xintercept = mean_change, color = group),
             linetype = "solid", size = 1) +
  scale_fill_manual(values = c("Observed" = "#2E86AB", "Imputed" = "#A23B72")) +
  scale_color_manual(values = c("Observed" = "#2E86AB", "Imputed" = "#A23B72")) +
  labs(
    title = "Distribution of Vision Changes (Simplified)",
    subtitle = "Vertical lines show group means",
    x = "Change in logMAR (positive = worsening)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# 2. Boxplots - detailed groups
p2_detailed <- ggplot(plot_data_detailed, aes(x = imputation_group, y = change_baseline_to_outcome, fill = imputation_group)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_fill_manual(values = c("Observed" = "#2E86AB", 
                              "Imputed (with intermediate)" = "#A23B72",
                              "Imputed (no intermediate)" = "#F18F01")) +
  labs(
    title = "Vision Changes by Detailed Group",
    subtitle = "Box shows IQR, whiskers extend to 1.5*IQR",
    y = "Change in logMAR (positive = worsening)",
    x = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# 3. Q-Q plot
observed_changes <- plot_data$change_baseline_to_outcome[plot_data$Group == "Observed"]
imputed_changes <- plot_data$change_baseline_to_outcome[plot_data$Group == "Imputed"]

# Sample to same size for Q-Q plot
n_min <- min(length(observed_changes), length(imputed_changes))
qq_data <- data.frame(
  observed = sort(sample(observed_changes, n_min)),
  imputed = sort(sample(imputed_changes, n_min))
)

p3 <- ggplot(qq_data, aes(x = observed, y = imputed)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(
    title = "Q-Q Plot: Imputed vs Observed Changes",
    subtitle = "Points should follow red line if distributions are similar",
    x = "Observed Change Quantiles",
    y = "Imputed Change Quantiles"
  ) +
  theme_minimal()

# 4. Cumulative distribution
p4 <- ggplot(plot_data, aes(x = change_baseline_to_outcome, color = Group)) +
  stat_ecdf(size = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = c("Observed" = "#2E86AB", "Imputed" = "#A23B72")) +
  labs(
    title = "Cumulative Distribution of Changes",
    x = "Change in logMAR",
    y = "Cumulative Probability"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Arrange plots - show detailed version first
grid.arrange(p1_detailed, p2_detailed, ncol = 2)

# Then show additional comparisons
grid.arrange(p3, p4, ncol = 2)
```

## Directional Change Analysis

```{r directional-changes}
# Compare proportions across detailed groups
direction_summary_detailed <- changes_df |>
  filter(!is.na(outcome_direction)) |>
  group_by(imputation_group, outcome_direction) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(imputation_group) |>
  mutate(
    total = sum(n),
    pct = round(100 * n / total, 1)
  )

# Reshape for display
direction_wide_detailed <- direction_summary_detailed |>
  select(imputation_group, outcome_direction, n, pct) |>
  pivot_wider(
    names_from = outcome_direction,
    values_from = c(n, pct),
    values_fill = list(n = 0, pct = 0)
  )

kable(direction_wide_detailed,
      caption = "Direction of vision change by detailed imputation group")

# Also create simplified version for backward compatibility
direction_summary <- changes_df |>
  filter(!is.na(outcome_direction)) |>
  group_by(outcome_was_imputed, outcome_direction) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(outcome_was_imputed) |>
  mutate(
    total = sum(n),
    pct = round(100 * n / total, 1),
    Group = ifelse(outcome_was_imputed, "Imputed", "Observed")
  )

# Reshape for display
direction_wide <- direction_summary |>
  select(Group, outcome_direction, n, pct) |>
  pivot_wider(
    names_from = outcome_direction,
    values_from = c(n, pct),
    values_fill = list(n = 0, pct = 0)
  )

kable(direction_wide,
      caption = "Direction of vision change by imputation status")

# Chi-square test for independence
direction_table <- table(changes_df$outcome_was_imputed, changes_df$outcome_direction)
chi_test <- chisq.test(direction_table)

chi_result <- data.frame(
  Test = "Chi-square test of independence",
  Chi_Square = chi_test$statistic,
  df = chi_test$parameter,
  P_Value = chi_test$p.value,
  Interpretation = ifelse(chi_test$p.value < 0.05, 
                         "Proportions differ between groups",
                         "Proportions similar between groups")
)

kable(chi_result, digits = 4)

# Visualization
direction_plot_data <- direction_summary |>
  mutate(outcome_direction = factor(outcome_direction, 
                                   levels = c("Improved", "Stable", "Worsened")))

p_direction <- ggplot(direction_plot_data, 
                     aes(x = Group, y = pct, fill = outcome_direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = paste0(pct, "%")), 
           position = position_dodge(width = 0.9),
           vjust = -0.5) +
  scale_fill_manual(values = c("Improved" = "#2ca02c", 
                              "Stable" = "#ff7f0e", 
                              "Worsened" = "#d62728")) +
  labs(
    title = "Direction of Vision Change by Group",
    y = "Percentage",
    x = "",
    fill = "Change Direction"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p_direction)
```

## Baseline Predictors of Change

This section examines whether baseline characteristics (like CCI) predict the magnitude of change, which might explain differences between groups.

```{r baseline-predictors}
# Check what variables are available in the original data
available_vars <- names(df_original)
baseline_vars <- intersect(c("row_id", "CCI", "DCSI", "age"), available_vars)

>>>>

---
title: "Reanalysis Step 3: Comprehensive Imputation Diagnostics"
subtitle: "Assessment of 2l.pan clustered multiple imputation quality"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 3
    fig-width: 10
    fig-height: 7
---

@@

This notebook provides comprehensive diagnostics for the 2l.pan clustered multiple imputation, assessing:

1. **Clustering structure** - Intra-patient correlations and bilateral eye relationships
2. **Cohort differences** - Submitted (observed outcomes) vs Excluded (imputed outcomes)  
3. **Imputation quality** - Distribution comparisons, temporal consistency, clinical plausibility
4. **Multiple imputation variability** - Between-imputation variance and convergence

@@

library(mice)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)

# Initialize logging
log_file <- init_log("imputation_diagnostics_combined")
```

## Data Loading

```{r load-data}
# Load original and imputed data
df_original <- readRDS(file.path(reanalysis_data_dir, "df_for_imputation.rds"))
imputed_datasets <- readRDS(file.path(reanalysis_data_dir, "imputed_datasets.rds"))
n_imputations <- length(imputed_datasets)

# Load mids object if available
if (file.exists(file.path(reanalysis_data_dir, "mids_object.rds"))) {
  mids_obj <- readRDS(file.path(reanalysis_data_dir, "mids_object.rds"))
} else {
  mids_obj <- NULL
}

# Key statistics
n_eyes <- nrow(df_original)
n_patients <- n_distinct(df_original$e_mrn_deidentified)
n_bilateral <- sum(table(df_original$e_mrn_deidentified) == 2)
n_missing_outcome <- sum(is.na(df_original$outcome_VA_logMAR))

cat(sprintf("Dataset: %d eyes from %d patients\n", n_eyes, n_patients))
cat(sprintf("Bilateral patients: %d (%.1f%%)\n", n_bilateral, 100*n_bilateral/n_patients))
cat(sprintf("Missing outcomes: %d eyes (%.1f%%)\n", n_missing_outcome, 100*n_missing_outcome/n_eyes))
cat(sprintf("Imputations performed: %d\n", n_imputations))
```

## 1. Cohort Structure Analysis

```{r cohort-structure}
# Cohort comparison
cohort_summary <- df_original |>
  group_by(cohort_id) |>
  summarise(
    n_eyes = n(),
    n_patients = n_distinct(e_mrn_deidentified),
    pct_of_total = round(100 * n_eyes / nrow(df_original), 1),

    # VA availability
    has_baseline = sum(!is.na(baseline_VA_logMAR)),
    has_intermediate = sum(!is.na(intermediate_VA_logMAR)),
    has_outcome = sum(!is.na(outcome_VA_logMAR)),

    # Clinical characteristics
    mean_age = round(mean(age, na.rm = TRUE), 1),
    pct_male = round(100 * mean(gender_cat == "Male", na.rm = TRUE), 1),
    mean_baseline_VA = round(mean(baseline_VA_logMAR, na.rm = TRUE), 3),

    # Key outcome: lapse rate
    pct_lapsed = round(100 * mean(ever_lapse_binary == 1, na.rm = TRUE), 1),

@@

kable(cohort_summary, 
      caption = "Cohort structure: Submitted (has 2-year outcome) vs Excluded (missing 2-year outcome)") |>
  kable_styling(full_width = FALSE)

# Missingness patterns
missing_patterns <- df_original |>
  mutate(
    pattern = case_when(
      !is.na(baseline_VA_logMAR) & !is.na(intermediate_VA_logMAR) & !is.na(outcome_VA_logMAR) ~ "Complete",
      !is.na(baseline_VA_logMAR) & !is.na(intermediate_VA_logMAR) & is.na(outcome_VA_logMAR) ~ "Missing outcome",
      !is.na(baseline_VA_logMAR) & is.na(intermediate_VA_logMAR) & is.na(outcome_VA_logMAR) ~ "Missing intermediate+outcome",
      TRUE ~ "Other"
    )
  ) |>
  group_by(cohort_id, pattern) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = pattern, values_from = n, values_fill = 0)

kable(missing_patterns, caption = "VA missingness patterns by cohort") |>
  kable_styling(full_width = FALSE)
```

## 2. Intra-Patient Correlation (Clustering Effect)

```{r clustering-analysis}
# Identify bilateral patients
bilateral_ids <- df_original |>
  count(e_mrn_deidentified) |>
  filter(n == 2) |>
  pull(e_mrn_deidentified)

# Calculate between-eye correlations
bilateral_data <- df_original |>
  filter(e_mrn_deidentified %in% bilateral_ids) |>
  group_by(e_mrn_deidentified) |>
  arrange(row_number()) |>
  mutate(eye = paste0("eye", row_number())) |>
  ungroup()

# Reshape for correlation
bilateral_wide <- bilateral_data |>
  select(e_mrn_deidentified, eye, baseline_VA_logMAR, intermediate_VA_logMAR, outcome_VA_logMAR) |>
  pivot_wider(names_from = eye, values_from = c(baseline_VA_logMAR, intermediate_VA_logMAR, outcome_VA_logMAR))

# Calculate observed correlations
cor_baseline <- cor(bilateral_wide$baseline_VA_logMAR_eye1, bilateral_wide$baseline_VA_logMAR_eye2, use = "complete.obs")
cor_intermediate <- cor(bilateral_wide$intermediate_VA_logMAR_eye1, bilateral_wide$intermediate_VA_logMAR_eye2, use = "complete.obs")
cor_outcome_obs <- cor(bilateral_wide$outcome_VA_logMAR_eye1, bilateral_wide$outcome_VA_logMAR_eye2, use = "complete.obs")

# Calculate imputed correlation (first imputation)
bilateral_imp <- imputed_datasets[[1]] |>
  filter(e_mrn_deidentified %in% bilateral_ids) |>
  group_by(e_mrn_deidentified) |>
  arrange(row_number()) |>
  mutate(eye = paste0("eye", row_number())) |>
  ungroup() |>
  select(e_mrn_deidentified, eye, outcome_VA_logMAR) |>
  pivot_wider(names_from = eye, values_from = outcome_VA_logMAR)

cor_outcome_imp <- cor(bilateral_imp$outcome_VA_logMAR_eye1, bilateral_imp$outcome_VA_logMAR_eye2, use = "complete.obs")

# Summary table
intra_patient_cors <- data.frame(
  Timepoint = c("Baseline", "Intermediate", "Outcome (Observed)", "Outcome (After Imputation)"),
  N_Pairs = c(
    sum(complete.cases(bilateral_wide[,c("baseline_VA_logMAR_eye1", "baseline_VA_logMAR_eye2")])),
    sum(complete.cases(bilateral_wide[,c("intermediate_VA_logMAR_eye1", "intermediate_VA_logMAR_eye2")])),
    sum(complete.cases(bilateral_wide[,c("outcome_VA_logMAR_eye1", "outcome_VA_logMAR_eye2")])),
    nrow(bilateral_imp)
  ),
  Correlation = round(c(cor_baseline, cor_intermediate, cor_outcome_obs, cor_outcome_imp), 3),
  Interpretation = c(
    "Natural between-eye correlation",
    "Correlation at intermediate timepoint", 
    "Observed outcomes only (limited n)",
    "All bilateral patients after 2l.pan imputation"
  )
)

kable(intra_patient_cors, 
      caption = paste0("Intra-patient (between-eye) correlations (", length(bilateral_ids), " bilateral patients)")) |>
  kable_styling(full_width = FALSE)

# Visualization
p_bilateral <- ggplot(bilateral_imp, aes(x = outcome_VA_logMAR_eye1, y = outcome_VA_logMAR_eye2)) +
  geom_point(alpha = 0.4, color = "#2E86AB") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = sprintf("Between-Eye Correlation After 2l.pan Imputation (r = %.3f)", cor_outcome_imp),
    x = "Eye 1 Outcome VA (logMAR)",
    y = "Eye 2 Outcome VA (logMAR)"
  ) +
  theme_minimal()

print(p_bilateral)
```

## 3. Distribution Comparison: Observed vs Imputed

```{r distribution-comparison}
# Add imputation indicators and calculate changes
changes_df <- imputed_datasets[[1]] |>
  mutate(
    outcome_was_imputed = df_original$cohort_id == "excluded",
    change_baseline_outcome = outcome_VA_logMAR - baseline_VA_logMAR,
    change_category = case_when(
      change_baseline_outcome < -0.1 ~ "Improved",
      change_baseline_outcome > 0.1 ~ "Worsened", 
      TRUE ~ "Stable"
    )
  )

# Summary statistics by imputation status
change_summary <- changes_df |>
  filter(!is.na(change_baseline_outcome)) |>
  group_by(outcome_was_imputed) |>
  summarise(
    n = n(),
    mean = round(mean(change_baseline_outcome), 3),
    sd = round(sd(change_baseline_outcome), 3),
    median = round(median(change_baseline_outcome), 3),
    q25 = round(quantile(change_baseline_outcome, 0.25), 3),
    q75 = round(quantile(change_baseline_outcome, 0.75), 3),
    .groups = "drop"
  ) |>
  mutate(Group = ifelse(outcome_was_imputed, "Imputed", "Observed"))

kable(change_summary |> select(Group, everything(), -outcome_was_imputed),
      caption = "Distribution of vision changes (outcome - baseline)") |>
  kable_styling(full_width = FALSE)

# Statistical tests
t_test <- t.test(change_baseline_outcome ~ outcome_was_imputed, data = changes_df)
ks_test <- ks.test(
  changes_df$change_baseline_outcome[changes_df$outcome_was_imputed],
  changes_df$change_baseline_outcome[!changes_df$outcome_was_imputed]
)

test_results <- data.frame(
  Test = c("Two-sample t-test", "Kolmogorov-Smirnov"),
  Statistic = round(c(t_test$statistic, ks_test$statistic), 3),
  P_Value = round(c(t_test$p.value, ks_test$p.value), 4),
  Interpretation = c(
    ifelse(t_test$p.value < 0.05, "Means differ significantly", "Means are similar"),
    ifelse(ks_test$p.value < 0.05, "Distributions differ", "Distributions are similar")
  )
)

kable(test_results, caption = "Statistical comparison of observed vs imputed changes") |>
  kable_styling(full_width = FALSE)

# Visualization
p1 <- ggplot(changes_df, aes(x = change_baseline_outcome, fill = factor(outcome_was_imputed))) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_fill_manual(values = c("FALSE" = "#2E86AB", "TRUE" = "#A23B72"),
                    labels = c("Observed", "Imputed")) +
  labs(title = "Distribution of Vision Changes",
       x = "Change in logMAR (positive = worsening)",
       y = "Density",
       fill = "Group") +
  theme_minimal() +
  theme(legend.position = "bottom")

p2 <- ggplot(changes_df, aes(x = factor(outcome_was_imputed), y = change_baseline_outcome, 
                             fill = factor(outcome_was_imputed))) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_fill_manual(values = c("FALSE" = "#2E86AB", "TRUE" = "#A23B72")) +
  scale_x_discrete(labels = c("Observed", "Imputed")) +
  labs(title = "Vision Changes by Group",
       y = "Change in logMAR",
       x = "") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```

## 4. Clinical Relationships: Lapse Effect

```{r lapse-analysis}
# Lapse-outcome relationship
lapse_summary <- changes_df |>
  group_by(cohort_id, ever_lapse_binary) |>
  summarise(
    n = n(),
    mean_outcome = round(mean(outcome_VA_logMAR, na.rm = TRUE), 3),
    mean_change = round(mean(change_baseline_outcome, na.rm = TRUE), 3),
    pct_VI = round(100 * mean(outcome_VA_logMAR >= 0.3, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  mutate(lapse_status = ifelse(ever_lapse_binary == 1, "Lapsed", "Not Lapsed")) |>
  select(-ever_lapse_binary) |>
  arrange(cohort_id, lapse_status)

kable(lapse_summary, 
      caption = "Vision outcomes by lapse status and cohort") |>
  kable_styling(full_width = FALSE)

# Effect sizes
effect_sizes <- lapse_summary |>
  group_by(cohort_id) |>
  summarise(
    diff_outcome = mean_outcome[lapse_status == "Lapsed"] - mean_outcome[lapse_status == "Not Lapsed"],
    diff_change = mean_change[lapse_status == "Lapsed"] - mean_change[lapse_status == "Not Lapsed"],
    diff_VI_pct = pct_VI[lapse_status == "Lapsed"] - pct_VI[lapse_status == "Not Lapsed"],
    .groups = "drop"
  ) |>
  mutate(
    clinical_interpretation = ifelse(diff_outcome > 0, 
                                    "Lapsed have worse vision (expected)", 
                                    "Lapsed have better vision (unexpected)")
  )

kable(effect_sizes,
      caption = "Effect of lapse on outcomes (positive = worse for lapsed)") |>
  kable_styling(full_width = FALSE)

# Visualization
p_lapse <- ggplot(changes_df, 
                  aes(x = factor(ever_lapse_binary, labels = c("Not Lapsed", "Lapsed")),
                      y = outcome_VA_logMAR,
                      fill = factor(ever_lapse_binary))) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ cohort_id) +
  scale_fill_manual(values = c("0" = "#2E86AB", "1" = "#A23B72")) +
  labs(title = "Outcome VA by Lapse Status and Cohort",
       y = "Outcome VA (logMAR)",
       x = "") +
  theme_minimal() +
  theme(legend.position = "none")

print(p_lapse)
```

## 5. Temporal Consistency

```{r temporal-consistency}
# For patients with intermediate values, check trajectory consistency
temporal_df <- imputed_datasets[[1]] |>
  filter(!is.na(intermediate_VA_logMAR)) |>
  mutate(
    change_1 = intermediate_VA_logMAR - baseline_VA_logMAR,
    change_2 = outcome_VA_logMAR - intermediate_VA_logMAR,
    total_change = outcome_VA_logMAR - baseline_VA_logMAR,

    # Check arithmetic consistency
    calculated_total = change_1 + change_2,
    discrepancy = abs(total_change - calculated_total),

    # Check direction consistency
    trajectory_reversal = sign(change_1) != sign(change_2) & abs(change_1) > 0.1 & abs(change_2) > 0.1
  )

temporal_summary <- temporal_df |>
  group_by(cohort_id) |>
  summarise(
    n = n(),
    mean_discrepancy = round(mean(discrepancy), 5),
    max_discrepancy = round(max(discrepancy), 5),
    pct_reversals = round(100 * mean(trajectory_reversal), 1),
    cor_segments = round(cor(change_1, change_2, use = "complete.obs"), 3),
    .groups = "drop"
  )

kable(temporal_summary,
      caption = "Temporal consistency for patients with intermediate values") |>
  kable_styling(full_width = FALSE)

# Sample trajectories
set.seed(2025)
sample_ids <- sample(temporal_df$row_number(), min(200, nrow(temporal_df)))

trajectory_plot_data <- temporal_df[sample_ids,] |>
  mutate(patient_id = row_number()) |>
  select(patient_id, cohort_id, baseline_VA_logMAR, intermediate_VA_logMAR, outcome_VA_logMAR) |>
  pivot_longer(cols = c(baseline_VA_logMAR, intermediate_VA_logMAR, outcome_VA_logMAR),
               names_to = "timepoint", values_to = "VA") |>
  mutate(timepoint = factor(timepoint, 
                           levels = c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR"),
                           labels = c("Baseline", "Intermediate", "2 Years")))

p_traj <- ggplot(trajectory_plot_data, 
                 aes(x = timepoint, y = VA, group = patient_id, color = cohort_id)) +
  geom_line(alpha = 0.3) +
  facet_wrap(~ cohort_id) +
  scale_color_manual(values = c("submitted" = "#2E86AB", "excluded" = "#A23B72")) +
  labs(title = "Sample Patient Trajectories (n=200 with intermediate values)",
       x = "", y = "VA (logMAR)") +
  theme_minimal() +
  theme(legend.position = "none")

print(p_traj)
```

## 6. Multiple Imputation Variance

```{r mi-variance}
# Calculate summary statistics across all imputations
mi_summaries <- lapply(1:n_imputations, function(i) {
  imputed_datasets[[i]] |>
    group_by(cohort_id) |>
    summarise(
      mean_outcome = mean(outcome_VA_logMAR, na.rm = TRUE),
      sd_outcome = sd(outcome_VA_logMAR, na.rm = TRUE),
      pct_VI = mean(outcome_VA_logMAR >= 0.3, na.rm = TRUE) * 100,
      .groups = "drop"
    ) |>
    mutate(imputation = i)
})

mi_combined <- bind_rows(mi_summaries)

# Calculate variance components (Rubin's rules)
mi_variance <- mi_combined |>
  group_by(cohort_id) |>
  summarise(
    # Pooled estimate
    pooled_mean = round(mean(mean_outcome), 3),

    # Within-imputation variance
    W = mean(sd_outcome^2),

    # Between-imputation variance  
    B = var(mean_outcome),
